Evaluating FAIR Digital Object and Linked Data as distributed object systems

### 2023-11-16 latex/main.tex
Typos, remove section crossreferences

%% Definitions needed for pandoc's silly output
\usepackage{hyperref}
\hypersetup{
  pdflang={en-GB},
  hidelinks
}
\usepackage{varioref}

@@

@@

\section*{Introduction}\label{sec:introduction}

@@

The rest of this paper is organised as follows: We begin with a background primer on FDO and Linked Data to provide a foundation for the work. In the Methods section, we introduce the conceptual frameworks we use for comparison. Subsequently, in the Results section, we systematically step through the outcomes of applying these frameworks to both FDO and Linked Data. For each framework, we derive key observations. We end with a discussion of these results and their implications for both approaches and conclude.

\section*{Background and related work}\label{sec:background}

@@

\subsection*{FAIR Digital Object}\label{sec:fdo}

@@

FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}. Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}. The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents.

\subsubsection*{FDO approaches}\label{sec:fdo-approaches}

@@

From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we will cover \vpageref{sec:ld}.

\subsubsection*{Next steps for FDO}\label{sec:next-step-fdo}

The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-06-17, most of these documents are open for public review, while some are still in draft stages for internal review.  We provide an overview of these documents in Appendix \ref{sec:appendixfdo}. These documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}.  Except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations. Going forward a key strategy of the Forum is the use of profiles to help define specific attributes in metadata that are necessary for domains or application contexts. However, these are not yet fully implemented in the implementations considered here.

@@

\subsection*{From the Semantic Web to Linked Data}\label{sec:ld}

@@

\subsubsection*{A brief history of the Semantic Web}\label{sec:semweb}

@@

\subsubsection*{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}

@@

\section*{Methods}\label{sec:methods}

@@

\section*{Results}\label{sec:results}

\subsection*{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}

@@

\subsubsection*{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}

@@

From this mapping\footnote{A SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{soilandreyes2023}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology. Notable are that neither DOIP or HTTP have strong support for transactions (explored further \vpageref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.

@@

\subsection*{Assessing FDO implementations}\label{sec:doip-fdo-compare}

@@

In Table \vref{tbl:fdo-checks}, we evaluate completeness of the guidelines in two current FDO realisations: 1) DOIPv2 \autocite{foundationDigitalObjectInterface} and 2) Linked Data Platform \autocite{w3-ldp}, as proposed by \textcite{FDOFramework}. We provide our analysis of each realization with respect to the FDO Guidline and also provide suggestions for that realization to meet the given guideline.

@@

\subsection*{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}

@@

\subsection*{Assessing FDO against FAIR}\label{sec:fair-compare}

@@

\subsection*{EOSC Interoperability Framework}\label{sec:eosc-interoperability-framework}

@@

As covered in our introduction\vref{sec:introduction}, EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts. Therefore we here find it important to summarize how FDO and Linked Data can help satisfy the EOSC requirements.

@@

\section*{Discussion}\label{sec:discussion}

@@

Having considered FDO and the Web architecture as interoperability frameworks (\vref*{sec:interoperability-compare}), we observe that neither are magic bullets, but each bring different aspects of interoperability. The Web comes with a large degree of flexibility and openness, however this means interoperability can suffer as services have different APIs and data models, although with common patterns. This is also true for Linked Data on the Web, with many overlapping ontologies and frequent inconsistencies in resolution mechanisms; although somewhat alleviated in recent years by schema.org becoming common metadata model for semantic markup inline in Web pages. The Web is based on a common HTTP protocol which has remained stable architecturally throughout its 32 years of largely backwards-compatible evolution. FDO on the other side sets down multiple rigid rules for identifiers, types, methods etc. that are advanterous for interoperability and predictability for FAIR consumption. Yet there is a large degree of freedom in how the FDO rules can be implemented by a given community, for instance there is no common metadata model or identifier resolution mechanism, and DOIP is just one possible transport method for FDOs, which itself does not enforce these rules.

When evaluating FDO implementations against the FDO guidelines (\vref*{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used. Achieving fully interoperable FAIR Digital Objects would require further convergence on implementation practices, and it is not given that this needs to diverge from the established Web architecture.  It is not clear from FDO guidelines if moving from HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects will benefit FAIR practitioners, when both approaches require additional equably implementable restrictions and conventions, such as using persistent identifiers or pre-defining an object's type.

Considering this, by comparing FDO and Web as middleware (\vref*{sec:middleware}) we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g. lack of concurrency and transparancy. Recent updates to HTTP have added many features needed for large-scale usage such as video streaming services (e.g. caching, multiplexing, cloud deployments), and having the option to transparently apply these also to FDOs seems like a strong incentive. Many programmatic features for distributed objects are however missing or needing custom extensions in both aspects, such as transactions, asynchronous operations and streaming.

By assessing FDO against the FAIR principles (\vref*{sec:fair-compare}) we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence). While there are implementations of each of these in general Linked Data examples, there is no single set of implementation guides that fully realizes the FAIR principles. \emph{FAIRification} efforts like the FAIR Cookbook \autocite{faircookbook} and FAIR Implementation Profiles \autocite{FIP} are bringing existing practices together, but there remains a potential role for FDO in giving a coherent set of implementation practices that can practically achieve FAIR. Significant effort, also within EOSC, is now moving towards FAIR metrics \autocite{Devaraju_2021}, which in practice need to make additional assumptions on how FAIR principles are implemented, but these are not always formalized \autocite{10.5281/zenodo.7463421} nor can they be taken to be universally correct \autocite{10.5281/zenodo.7848102}. Given that most of the existing FAIR guides and assessment tools are focused on Web and Linked Data, it would be reasonable for FDO to then provide a profile of such implementation choices that can achieve best of both worlds.

EOSC has been largely supportive of FDO, FAIR and related services. By contrasting the EOSC Interoperability Framework (\vref*{sec:eosc-interoperability-framework}) with FDO, we found that there are important dimensions that are not solved at a technical level, but through organization collaboration, legal requirements and building community practices. FDO recommendations highlight community aspects, but at the same time the largest FAIR communities in many science domains are already producing and consuming Linked Data. Just as the Linked Data community has a challenge in convincing more research fields to use Semantic Web technologies, FDO currently need to build many new communities in areas that have shown interest in that approach (e.g. material science).  It may be advantageous for both these effort to be aligned and jointly promoted under the EOSC umbrella.

@@

\subsection*{What does FDO mean for Linked Data?}\label{sec:what-does-it-mean-for-linked-data}

@@

\section*{Conclusion}\label{sec:conclusion}

@@

%\clearpage
\defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}
\printshorthands[heading=fdobibliography]

>>>>

%% Definitions needed for pandoc's silly output

@@

%\usepackage[usetoc]{titleref}
\usepackage{hyperref}
\hypersetup{
  pdflang={en-GB},
  hidelinks
  }

%\usepackage{varioref}
\newcommand{\vref}{\ref}
\newcommand{\vpageref}{\ref}

@@

\section*{Introduction}
\label{sec:introduction}

@@

The rest of this paper is organised as follows: We begin with a background primer on FDO and Linked Data to provide a foundation for the work. In the Method section, we introduce the conceptual frameworks we use for comparison. Subsequently, in the Results section, we systematically step through the outcomes of applying these frameworks to both FDO and Linked Data. For each framework, we derive key observations. We end with a discussion of these results and their implications for both approaches and conclude.

\section*{Background and related work}
\label{sec:background}

@@

\subsection*{FAIR Digital Object}
\label{sec:fdo}

@@

FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}. Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}. The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents \autocite{fdo-Specs}.

\subsubsection*{FDO approaches}
\label{sec:fdo-approaches}

@@

From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we will cover in section \emph{From the Semantic Web to Linked Data}.
%\vref{sec:ld}

\subsubsection*{Next steps for FDO}
\label{sec:next-step-fdo}

The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed \href{https://fairdo.org/specifications/}{requirement documents} \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-06-17, most of these documents are open for public review, while some are still in draft stages for internal review.  We provide an overview of these documents in Appendix \ref{sec:appendixfdo}. These documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}.  Except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations. Going forward a key strategy of the Forum is the use of profiles to help define specific attributes in metadata that are necessary for domains or application contexts. However, these are not yet fully implemented in the implementations considered here.

@@

\subsection*{From the Semantic Web to Linked Data}
\addcontentsline{toc}{subsection}{From the Semantic Web to Linked Data}
\label{sec:ld}

@@

\subsubsection*{A brief history of the Semantic Web}
\label{sec:semweb}

@@

\subsubsection*{Linked Data: Rebuilding the Web of Data}
\label{sec:ld-web}

@@

\section*{Method}
\label{sec:methods}

@@

\section*{Results}
\label{sec:results}

\subsection*{Considering FDO/Web as interoperability framework for Fast Data}
\label{sec:interoperability-compare}

@@

\subsubsection*{Mapping of Metamodel concepts}
\label{mapping-of-metamodel-concepts}

@@

From this mapping\footnote{A SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{soilandreyes2023}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology. Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \emph{Comparing FDO and Web as middleware infrastructures}),
%\vpageref{sec:middleware}
as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.

@@

\subsection*{Assessing FDO implementations}
\label{sec:doip-fdo-compare}

@@

In Table \vref{tbl:fdo-checks}, we evaluate completeness of the guidelines in two current FDO realisations: 1) DOIPv2 \autocite{foundationDigitalObjectInterface} and 2) Linked Data Platform \autocite{w3-ldp}, as proposed by \textcite{FDOFramework}. We provide our analysis of each realisation with respect to the FDO Guideline and also provide suggestions for that realisation to meet the given guideline.

@@

\subsection*{Comparing FDO and Web as middleware infrastructures}
\label{sec:middleware}

@@

\subsection*{Assessing FDO against FAIR}
\label{sec:fair-compare}

@@

\subsection*{EOSC Interoperability Framework}
\label{sec:eosc-interoperability-framework}

@@

As covered in our introduction,
%\vref{sec:introduction}, 
EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts. Therefore we here find it important to summarize how FDO and Linked Data can help satisfy the EOSC requirements.

@@

\section*{Discussion}
\label{sec:discussion}

@@

Having considered FDO and the Web architecture as interoperability frameworks 
%(\vref*{sec:interoperability-compare})
we observe that neither are magic bullets, but each bring different aspects of interoperability. The Web comes with a large degree of flexibility and openness, however this means interoperability can suffer as services have different APIs and data models, although with common patterns. This is also true for Linked Data on the Web, with many overlapping ontologies and frequent inconsistencies in resolution mechanisms; although somewhat alleviated in recent years by schema.org becoming common metadata model for semantic markup inline in Web pages. The Web is based on a common HTTP protocol which has remained stable architecturally throughout its 32 years of largely backwards-compatible evolution. FDO on the other side sets down multiple rigid rules for identifiers, types, methods etc. that are advanterous for interoperability and predictability for FAIR consumption. Yet there is a large degree of freedom in how the FDO rules can be implemented by a given community, for instance there is no common metadata model or identifier resolution mechanism, and DOIP is just one possible transport method for FDOs, which itself does not enforce these rules.

When evaluating FDO implementations against the FDO guidelines 
%(\vref*{sec:doip-fdo-compare}) 
we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used. Achieving fully interoperable FAIR Digital Objects would require further convergence on implementation practices, and it is not given that this needs to diverge from the established Web architecture.  It is not clear from FDO guidelines if moving from HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects will benefit FAIR practitioners, when both approaches require additional equably implementable restrictions and conventions, such as using persistent identifiers or pre-defining an object's type.

Considering this, by comparing FDO and Web as middleware 
%(\vref*{sec:middleware}) 
we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g. lack of concurrency and transparancy. Recent updates to HTTP have added many features needed for large-scale usage such as video streaming services (e.g. caching, multiplexing, cloud deployments), and having the option to transparently apply these also to FDOs seems like a strong incentive. Many programmatic features for distributed objects are however missing or needing custom extensions in both aspects, such as transactions, asynchronous operations and streaming.

By assessing FDO against the FAIR principles 
%(\vref*{sec:fair-compare}) 
we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence). While there are implementations of each of these in general Linked Data examples, there is no single set of implementation guides that fully realizes the FAIR principles. \emph{FAIRification} efforts like the FAIR Cookbook \autocite{faircookbook} and FAIR Implementation Profiles \autocite{FIP} are bringing existing practices together, but there remains a potential role for FDO in giving a coherent set of implementation practices that can practically achieve FAIR. Significant effort, also within EOSC, is now moving towards FAIR metrics \autocite{Devaraju_2021}, which in practice need to make additional assumptions on how FAIR principles are implemented, but these are not always formalized \autocite{10.5281/zenodo.7463421} nor can they be taken to be universally correct \autocite{10.5281/zenodo.7848102}. Given that most of the existing FAIR guides and assessment tools are focused on Web and Linked Data, it would be reasonable for FDO to then provide a profile of such implementation choices that can achieve best of both worlds.

EOSC has been largely supportive of FDO, FAIR and related services. By contrasting the EOSC Interoperability Framework 
%(\vref*{sec:eosc-interoperability-framework})
with FDO, we found that there are important dimensions that are not solved at a technical level, but through organization collaboration, legal requirements and building community practices. FDO recommendations highlight community aspects, but at the same time the largest FAIR communities in many science domains are already producing and consuming Linked Data. Just as the Linked Data community has a challenge in convincing more research fields to use Semantic Web technologies, FDO currently need to build many new communities in areas that have shown interest in that approach (e.g. material science).  It may be advantageous for both these effort to be aligned and jointly promoted under the EOSC umbrella.

@@

\subsection*{What does FDO mean for Linked Data?}
\label{sec:what-does-it-mean-for-linked-data}

@@

\section*{Conclusion}
\label{sec:conclusion}

@@

%\clearpage
\defbibheading{fdobibliography}{\subsection*{FDO Specifications}
\label{sec:fdo-bibliography}}
\printshorthands[heading=fdobibliography]
